---
title: "calculus1_eaquinto"
---

This is a Quarto website.

To learn more about Quarto websites visit <https://quarto.org/docs/websites>.


Intresting paradigm. Employers want experience. Graduate schools want grades. Competing interests... But they both ultimately want money. 


What's the point of education? Portfolio projects or mastery of subject? 

<br><br>

An Open Letter to Graduate Schools: 
----
- Dash dashboards
- Quarto Websites 
- coding word documents
- writing r packages to do calculus
- Quarto Revealjs for presentations (free, animations, speaker timer, speaker notes...)
- character & chapter summarising in readings


```{r}
library(tidyverse)

```

```{r}
fun.1 <- function(x) x^2 + x

p <- ggplot(data = data.frame(x = 0), mapping = aes(x = x))
```


```{r}
p + stat_function(fun = fun.1) + xlim(-5,5)
```

```{r}
p + layer(geom = "path",        # Default. Can be omitted.
          stat = "function",
          fun = fun.1,          # Give function
          mapping = aes(color = "fun.1") # Give a meaningful name to color
          ) +
    scale_x_continuous(limits = c(-5,5)) +
    scale_color_manual(name = "Function", values = c("blue"))
```

```{r}
fun.2 <- function(x) -1 * x + 10
fun.3 <- function(x) 3 * x + 2
```

```{r}
ggplot(data = data.frame(x = 0), mapping = aes(x = x))+ 
    layer(stat = "function",
          fun = fun.1,
          mapping = aes(color = fun.1)
          ) +
    layer(stat = "function",
          fun = fun.2,
          mapping = aes(color = fun.2)
          ) +
    layer(stat = "function",
          fun = fun.3,
          mapping = aes(color = fun.3)
          ) 
    # scale_x_continuous(limits = c(-5, 5)) +
    # scale_color_manual(name = "Functions",
    #                    values = c("blue", "red", "green"), # Color specification
    #                    labels = c("X^2 + X", "-X + 10", "3X + 2"))
```



```{r}
x <- 1:10
df <- data.frame(x)
ggplot(df, aes(x))+
    stat_function(fun=function(x) log(x)/(x-3))
```
